---
title: 'Lab 1: Intro to R, Git, and Github'
author: "ES _____"
date: 
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
  word_document: default
---



# Overview
In this we will focus on setting up R, RStudio, Git, Github,  and Sourcetree. 

R is a command-line interface that forces you to program your own analyses in a script. While this can be initimidating at first, it pays quickly to learn how to script your analyses when you have to redo your analysis if you discover a mistake, provide documentation, transparency and reproducibility of analyses, perform similar workflows, or perform batch prcessing. 

RStudio is a free and open-source integrated development environment for R, a programming language for statistical computing and graphics.

Git and Github are version control systems for tracking changes in computer files and coordinating work on those files among multiple people.

SourceTree is a Git GUI that offers a visual representation of your repositories.

## How These Tutorials Work
Because we are working with open source software, there are far more useful and varied examples online than we can possibly cover in this class. Part of learning open source software is by reading different turorials, blog posts, and other resources. I will post required and suggested reading, but you should consider these as springboards only. With a little bit of internet research, you can acheive far more than what these basic tutorials cover.

Sometimes I will give you instruction and examples in these tutorials, and sometimes I will point you to other online tutorials that you will have to complete. Note the learning objectives for each tutorial. These will be part of the overall course objectives, and will be covered in exams. 

## Lab Assignments
For each lab assignment I will ask you to demonstrate that you have done different pieces of the work, by prodviding maps or screen shots of images, plots or statistics calcuated from images, and some basic short writing and answering of questions. 
In each section, if the text is

><span style="color:CRIMSON;">CRIMSON like
this it indicates a prompt for a question that you *must* answer or a product you must turn in for your lab assignment. Crimson blocks of text are action items for you!</span> 

# Objectives  
After this lab, you will:  

* Have the majority of software for this class installed and ready to use
* Know how to utilize features of Git and Github manage version control of your code and share it.
* Know the basics of R and RStudio

## Helpful Terminology

Github has a glossary that will help in learning the lingo. It can be found here: https://help.github.com/articles/github-glossary/


## R
You can install the most recent version of R from here:
<https://cran.r-project.org/>

## RStudio
RStudio is one of several integrated development environments (IDEs), which provide convenient editors specialized in writing and executing R code. While it is not necessary to use R with an IDE, it makes scripting a lot easier. Consider this analogy: It is entirely possible to write your thesis in a basic text editor like Notepad or Microsoft Wordpad. However, it is much easier to write your thesis in a software like Microsoft Word, whihc provides functions like being able to control your page margins, font type and style, and providing basic spelling and grammar checks on your writing. Using an IDE for R provides similar functionality, plus much more, such as debugging, version control, webpage publication, and more (for example, these .html tutorials were all produced using RStudio).

Another free option is Eclipse <http://www.eclipse.org>, to be used with the Eclipse-based StatET plug-in <http://www.walware.de/goto/statet>. Eclipse allows you to semalessly move between multiple different scripting languages such as Java, Python, and R. 

For the purposes of this class, we will use RStudio. 

#Introduction to R
Launch RStudio. When you launch the IDE, it will launch an instance of R. The upper left panel shows the script editor where you write your code, with the command line below and on the right are several tabs that list your files, plots, R help packages and the package manager. Code from the script editor can be sent to R using the *Run* button or by pressing CRTL+ENTER. Only the current line or highlighted parts are sent to R to be executed. 

We want to use R to create code and a workflow that is more reproducible. With scripting, we can document everything we do! In science, our end goal isn't just to analyze stuff, but to do it in a way that anyone can easily and precisely replicate our workflow and results-this includes oursleves in 3 months when our advisor asks us to check something, or our paper reviews come back!

##Your First R Session
In R's console window, you will type next to the **>**. Use the mouse to click there and hit ENTER.
Type the following command:
```{r}
x<-c(1,2,4)
```
`x` is a variable, or in R parlance, an **object**. `<-` is the **assignment** operator. Object names cannot contain spaces. The *result* of the operation on the right hand side of `<-` is **assigned** to an **object** with the name specified on the left hand side of the`<-`. The result can be any type of R object, including your own functions (to come later!)

`c()` means "concatenate", i.e. it merges the numbers into a single **vector**. The parentheses, in this context, refer to a **function**.  The function name is "c" and the inputs are the things that are inside of the parentheses.

The `#` sign you see here means 'comment'.  Anything after it is ignored by R. The `##` sign shows a result from the R console.

One of the great things about R is that you can print any variable by just typing it and hitting enter.  
Do this now:
```{r}
x
```
This is very similar to the `print()` statement:
```{r}
print(x)
```
You should see:
`[1] 1 2 4`
Individual vector elements are accessed using `[ ]`, where [1] would be the first element, `[2]` the second, and so on...  Type:
```{r}
x[3]
```
This is what is known as a subscript" or **index**.  We can also do a subset. Type:
```{r}
x[2:3]
```
This asks R to print out the 2nd through the 3rd value in `x`. `:` is known as the **colon operator** and is also a function.

><span style="color:CRIMSON;">3. What happens if your try to print out the 4th element of x? Print your result and provide the answer in your lab write up.</span>

We can concatenate vectors into a new, larger vector.  Try:
```{r, results='hide'}
q <- c(x,x,8)
```

Let's do some simple statistics like calculating the mean and standard deviation of x.  Type:
```{r}
# Mean of the x vector
mean(x)
# Standard deviation of the x vector
sd(x) 
```
Notice these commands print the values out.  When you do an assignment `x <- ...` R will NOT print the values out.
But the assignment allows us to preserve the output as a new variable. Try:
```{r}
y <- mean(x)
# And then print it by:
y
```

><span style="color:CRIMSON;">4. Try creating a variable "s" that is the standard deviation of q. Make sure you print it out to confirm it worked.</span>  

##Some Basic Operations in R

You can get ar input by simply typing in some mathematical statements to evaluate:
```{r}
1+1
24/12
100^2
```
You can also make R recognize words. If you want R to recognize something you type as text, you must enclose it in quotes.
```{r}
#Make R write some words
paste("Remote","Sensing","is","covered","in","awesomesauce!")
```
Here, the `paste()` function works similarly to the concantenate function `c()`. Paste converts the arguments (the words in quotes) to character strings and then concantenates them. 

><span style="color:CRIMSON;">5. Can you make R write your name?.</span>

Of course, R is an object based language, so that means we can perform operations on objects we have assigned values to
```{r}
#Assign R objects
a <- 1+1
b <- 24/12
c <- 100^2
#Perform some math on the objects
d=(a+c)/b
#Print the results
d
```
*Typing Tip: If you are using RStudio, you can use a keyboard shortcut for the assignment operator.*
Windows/Linux: "Alt" + "-" or in Mac: "Option" + "-"

##Code & Comments in R
Everything you type into an R script is a code unless you tell it otherwise. Anything to the right of **#** is ignored by R. Use these comments within the code to describe what it is that you are coding. A good rule of thumb is to always comment liberally in your scripts. This will help you when you return to it and will also help other understand your scripts and analyses. 

Comments are important, and part of your grade will based on clear commenting of your code, so use it frequently. On the same line, you can include a comment, remember everything AFTER the pound sign will be ignored, so:
```{r}
#print out y
print(y)
```
Will do the same thing as:
```{r}
y
```
But the former line lets me know, in English, what it is doing.  

## List All Objects in the Environment  
Some functions are the same as in other languages. These might be familiar to you if you have ever worked with command line. Use`ls()`: to list objects in your current environment. Use `rm()`: remove objects from your current environment.

Give them a try:
```{r}
# assign value 100 to object "m"
m <- 100
ls()
#remove m
rm(m)
```

><span style="color:CRIMSON;">6. What objects are left in your R session after removing `m`?<span/>

Now try removing all of your objects
```{r}
#remove all objects
rm(list=ls())
ls()
```
using `rm(list=ls())`, you combine several functions to remove all objects. If you type `x` on the console now you should get `Error: object 'x' not found`. 

##A Little Deeper into R
Many R "packages" contain sample datasets, which we can list by:`data()`
Scroll through the list to see what sample datasets are available.  
Let's take a look at the Nile dataset. Print the Nile dataset by typing:
```{r}
Nile
```
We can plot this dataset by:
```{r}
plot(Nile)
```

This is a "time series" dataset (don't worry about what that means for now), but we can do basic stats on it:
```{r}
meanNile <- mean(Nile)
meanNile
sdNile <- sd(Nile)
sdNile
```
R has advanced plotting and statistical analysis, so let's create a histogram of this dataset:
```{r}
hist(Nile)
```

##Quitting an R session
Now let's quit R "properly" -- this command works on any OS:

`q()`

You will be prompted to `Save workspace image? [y/n/c]:`
Select no by typing `n` or click it.  
If you click `yes` (or type `y`) when you next open R, all of your datasets will re-load.

##Introduction to functions
Functions are the core of R -- in the previous step, we used multiple functions, including `c()`, `mean()`, `sd()`, `data()`, `hist()` and even `q()`  

Functions will always have parentheses which are used to pass parameters (some functions don't have parameters or don't need them, like `q()`.

Using and *writing* functions is a key part of learning any programming language. We'll do a quick introduction here.

Boot R back up again.  We didn't save our variables, so try:
```{r, results="hide"}
meanNile
```
You should see: `Error: object 'x' not found`

Now we are going to define a function.With all functions, we need to consider the following:
*What are the inputs?
*How does the function manipulate the inputs?
*What are the outputs that will be returned?

We are going to define a function that does the following:
*Input: a vector of values
*Output: a count of the number of odd values in the vector

Type the following in CAREFULLY.  Hit enter at the end of each line.A couple of things to note:

*The brackets demarcate the beginning `{` and end `}` of a function or operator 

*You will notice that until you finish typing the function in you will see a `+` instead of a `>`. This + is what is called a "line continuation character" and is a reminder that you haven't finished the function (with the end brace). 

```{r}
oddcount <- function(x) 
{
	k <- 0 # assign 0 to k
	for (n in x) {
		if(n %% 2 == 1)
		{
			k <- k+1 # %% is the modulo operator
		}
	}
  return(k)
}
```
If you get any errors, make sure you have a `>` and re-type the function until you don't get an error. Very common errors are not closing parentheses and braces. We will use tools later on to help with that.

Use the tab key to create the spacing. Tabs are not needed, but are key parts of keeping code readable.

Now we have defined the function oddcount. Notice that the **header** of the function tells us there is one input, `x`.  Recall that the input must be a vector, which we can create using `c()`. If you want to print the code of the function again, just type:
```{r}
# don't use parentheses!
oddcount
```
Now, let's try the function out by passing it a parameter:
```{r}
oddcount(x <- c(1,3,5))
```
There are three odd numbers in this vector.
Try:
```{r}
oddcount(x <- c(1,2,3,7,9))
```

><span style="color:CRIMSON;">7. How many odd numbers were in this vector?.</span> 

The function appears to be working, now let's look more carefully at what it is doing.

Following the *flow* of a function is a critical skill for understanding what it is doing (and fixing problems that may arise).

First, let's look at the modulo operator "%%".  This operator returns the remainder of a division, e.g.:
```{r}
38 %% 7
```
Notice that if the modulo of an even number divided by 2 is 0:
```{r}
38 %% 2
```
And an odd number is 1:
```{r}
39 %% 2
```
We'll learn more about for loops later on, but for now, `for(n in x) { ... }` 
can be interpreted as setting `n` equal to `x[1]`, running the commands inside of the braces,and then once it is done, repeating the process on `x[2]`, and so-on until it gets to the end of `x`.We can see this by typing:
```{r}
y <- c(3,0,7)
for(n in y) { print(n) } # Print simply prints the value of the variable
```
This is the same as:
```{r}
n <- y[1]
print(n)
n <- y[2]
print(n)
n <- y[3]
print(n)
```
Anyways, back to our oddcount function:
The function starts with k set to 0, and then starts the loop:
`for (n in x) {
if(n %% 2 == 1) k <- k+1 # %% is the modulo operator
}`
`x` was passed to the function as e.g.: `oddcount(x <- c(1,3,5))`, so `x` is `c(1,3,5)`
The loop starts with `n <- x[1]`, then tests whether or not `n` is odd by checking if `n %% 2` is equal (`==`, note the TWO equal signs) to 1.

Logical statements are used frequently, so let's see what `if()` is actually testing by typing:
```{r, results='hide'}
37 %% 2
37 %% 2 == 1  
38 %% 2
38 %% 2 == 1
```
It appears that an `if()` statement, fundamentally, is just testing whether a statement is `TRUE` or `FALSE`.  

So, if `n %% 2` *does* equal 1, the function adds `1` to `k`, and repeats the process on `x[2]`, and so on.

Once the function is done, we want to return the value of `k` which is keeping track of how many odd numbers are in our vector, so we use the `return()` statement:`return(k)`

Some definitions:
In the function definition, `x` is the "formal argument" or **formal parameter** of function oddcount
`c(1,3,5)` is the **actual argument** of the function.
Another note: functions can only return a SINGLE variable.  If we want to return a lot of things,we'll use some tricks discussed later.

A lot of times it is helpful to see, more specifically, what a function is doing by using `print()` statements.`paste()` merges the contents together into a single string.
We will modify the code a bit to help watch the function run:
```{r}
oddcount <- function(x) {
#	print("x is:")
	print(x)
	k <- 0 # assign 0 to k
	print(paste("k is initialized as",k))
	for (n in x) {
		print(paste("current x value being tested is",n))
		if(n %% 2 == 1) 
		{
			k <- k+1 # %% is the modulo operator
			print(paste(n,"is an odd number!"))
		} else
		{
			print(paste(n,"is an even number!"))
		}
		print(paste("k is currently",k))
	}
	print(paste("The final k is",k))
	return(k)
}
# And trying running our more verbose function:
oddcount(x <- c(1,2,3,7,9))
```

><span style="color:CRIMSON;">8. Try creating a new function "evencount" that counts the even numbers in a vector. Turn in your script as a .R script with appropriate comments.</span>  

# Directory Structures
Having a logical, consistent directory structure is an important component of every project inside this class, your other classes, your thesis, and life beyond your thesis. Again, when we do science, we are not just doing stuff, we are performing reproducible analyses. You need to layout your work, save different steps of your work, and be able to find it again in logical places. 

I will not dictate to you how to set up your directory structures for this class. Everyone's brain works a little bit differently when it comes to logical organization. Sometimes if you work on large, collaborative projects, they WILL dictate these structures. Fr good reason (again, science is about documentation, reproducibility, and sharing). For the purposes of this class, find something that works for you and stick with it. 

## Making good directory layouts
Before you set something up, read the following blog posts (yes, this is part of your homework!)

* On good directory layout: <https://nicercode.github.io/blog/2013-04-05-projects/>  
* On what RStudio offers for projects and version control: <https://support.rstudio.com/hc/en-us/articles/200526207-Using-Projects>

><span style="color:CRIMSON;">9. What are the three main reasons you want to have a good project layout? Can you think of any others?.</span>

><span style="color:CRIMSON;">10. What are the three primary principles to follow in a good project layout?</span>

## Full, base, and relative paths

The **full path** is the complete directions for how to find the desired directory or file. It *awlways* starts with the home directory or root (e.g., `C:\Users\DavidBowie\`). A full path is the **base path** when used to set the working directory to a specific directory. So in this class, David Bowie set up his directory layout so that his **full path** looks something like this: `C:\Users\DavidBowie\classes\mea591\Rscripts\lab2`. If David Bowie wanted to set a smart **base path** for his working directory (see below), it might look something like this: `C:\Users\DavidBowie\classes\mea591`.

*Note!* File or directory paths and the home directory look different on different operating systems. Linux will look something like this: `/home/DavidBowie`. Windows will look like the examples above: `C:\Documents and Settings\DavidBowie` or `C:\Users\DavidBowie`. Mac OS X and Unix look like this: `/Users/DavidBowie/`. The format will vary depending on what version of windows you are running. *NOTE* the direction of the slashes. I use a Mac, so get used to seeing examples following that structure, and make sure you adjust your paths according to your own operating system. 

><span style="color:CRIMSON;">11. Write out the full path for your R installation. Use the format of the operating system you are currently using.</span>

><span style="color:CRIMSON;">12. Write the path above using a different operating system.</span>

A relative path is a path to a  directory or file that starts from the location determined by the working directory. So, if our working directory is set to `/Users/DavidBowie/classes/mea591`, we can create a relative path for all directories and files within the `mea591` directory. 

><span style="color:CRIMSON;">13. Write out the **full path** for the directory structure you have set up for this class all the way to where you have saved this .html tutorial.</span>

><span style="color:CRIMSON;">14. Write out the **relative path** for this .html file assuming your working directory to be set to your equivalent of `Users/DavidBowie/classes/`.</span>

><span style="color:CRIMSON;">15. Write out the paths in an operating system other than your own.</span>

## The R working directory

In `R`, the working directory is the directory where are will look for any file to open and where it save any output. 

Without a working directory, all of your `R` scripts would need a full file path written out every time you want to open or save a file. It is much more efficient to set a **base file path** as our **working directory** and then all the file paths in our scripts are just the file path relative to that base path (this is known as a **relative path**).

### Find a full path to a file in an unknown location

Oh no! What if you can't figure out the file path to this .html file you downloaded? 

* Windows: Right click on file, select Properties, go to the General tab, or in the file path bar at the top of each navigator window.
* Mac OS X: Right click or CRTL+click on the file, select **Get Info**

### Determine current working directory

Once we are in `R`, we can view the current working directory using `getwd()`
```{r}
getwd()
```

### Set the working directory

To set our current working directory to the location where our data are located, we can use `R` script, or we can use the RStudio GUI. 

```{r}
#setwd('/Users/ehestir')
getwd()
```

To view the file contents and cub directories of your working directory:

```{r}
list.files()
```

You can also set the working directory using the RStudio GUI by navigating to `Session` > `Select Working Directory` > `Choose Directory`. 

Note that using either method, you will be able to see the ocntents of the current working directory in the Files pane of the RStudio GUI (to your right).

#Introduction to Git and Github

A great tutorial for using Git and Github with R can be found here http://happygitwithr.com/ 

Work through The Happy Git with R tutorial, ensuring you complete the steps below.


><span style="color:CRIMSON;">16. Complete Steps 1-28 in the "Happy Git with R" tutorial.</span>

><span style="color:CRIMSON;">17. In step 9 install SourceTree. Note: If you're using google chrome with a google account, and also use a google account for registering with Atlassian it makes life easy. You will not need to worry about setting up HTTPS or SSH authorization to make push and pull requests. </span>

><span style="color:CRIMSON;">18. Make a repo on GitHub called "____hw1. Write the render-read R script in 20.2 and commit it to your repo.</span>

><span style="color:CRIMSON;">19. Following the instructions in 23, clone a GitHub repository that interests you. Find one script in the repo and run the code locally and understand what it is doing. Once you understand what it is doing, document this in a R Notebook using RMarkdown and commit it to your hw1 repo. Make sure to cite the source of your script, and document any changes you made. DO NOT fork this code (we will do that next time).</span>

><span style="color:CRIMSON;">20. Following the instructions in 25, fork the 'bingo' repo. Clone it to your local machine and create a new bingo card. Commit and push your changes back to your copy of the repo on GitHub. Make a pull request back to the main 'bingo' repo.</span>

><span style="color:CRIMSON;">21. Following the instructions in 26, Burn it all down.</span>
